---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
load("movies.Rdata")
```



* * *

## Part 1: Data


* * *

## Part 2: Research question



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.
```{r audience-score}
summary(movies$audience_score)
hist(movies$audience_score)
```
```{r audience-rating}
summary(movies$audience_rating)
qplot(movies$audience_rating, geom="bar")
```

```{r imdb-rating}
summary(movies$imdb_rating)
hist(movies$imdb_rating)
```

```{r critics-rating}
ratings <- movies$critics_rating
summary(ratings)
qplot(ratings, geom="bar")
```

```{r critics-score}
summary(movies$critics_score)
```


* * *

## Part 4: Modeling

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 5: Prediction

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 6: Conclusion

